An Accurate Model for SARS-CoV-2 pooled RT-PCR Test Errors

### 2021-03-22 comment.tex
B4 merge
new fig

Studies focused on pooling for SARS-CoV-2 are in a consensus that sample dilution 
effects \cite{Dilution1, Dilution2} are not a concern \cite{Lion, DorfmanYuvalDor}.

Previous studies assumed that the probability of a true-positive (the
test's \emph{sensitivity}) does not depend on the number of infected
samples in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test
has been (assumed) identical for a pool with one sample from an
infected individual and, e.g., five such samples. This assumption is
common in the group testing literature \cite{Kim, OptimalDorfmanPool},
as well as in more specific, COVID-19 focused studies
\cite{Simplistic1, Simplistic2}. In this study we challenge these
commonly made assumptions and show how using a more accurate
probabilistic model affects estimation of false-negative rates for
Dorfman pooling.

@@

\subsection{Application}
As an application, we calculate the false-negative rate for a single \emph{infected} individual,
henceforth referred to as "Donald" under our model and under the common assumption. We distinguish three types of
false-negative events when performing pooling. A \emph{single test}'s
false-negative is the event of a negative result upon testing Donald
separately, i.e., in an RT-PCR test without pooling. A \emph{pooled}
false-negative occurs when a pooled test containing Donald's sample
(and other samples) yields a negative result, i.e., the pooling fails
to detect at least one positive result. Lastly, a \emph{scheme}
false-negative occurs when an entire pooling scheme fails to identify
Donald as infected. Our first task is to calculate Dorfman's scheme
false-negative rate. Rephrasing, we wish to answer the following
question: what is the probability of not identifying Donald as
infected under a Dorfman pooling scheme?

\subsubsection{Scheme false-negative rate}
Denote the prevalence of infection in the (tested) population $q$. We
assign Donald as individual $1$, so that $\theta_1=1$. Then:

@@

\subsubsection{Comparison metric}
The single test false-negative rate $1-\Se$ and scheme false-negative
rate $\mi$ are compared via:
\begin{equation}\label{eq:erel}
\rel{ours} := \frac{\mi - (1-\Se)}{1-\Se} \cdot 100.
\end{equation}
$\rel{ours}$ is the percentage increase in the pooling scheme false-negative
rate, relative to the single test false-negative rate.

According to the common assumption, the scheme false-negative rate is
$1-\Se^2$. A short calculation shows that this approximation implies
the percentage increase in scheme false-negative rate is just
$\rel{common} := 100\cdot \Se$.

\section{Results}\label{section:results}
We plot $\rel{ours}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\rel{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, an increase of at least $60\%$ in
$\rel{ours}$ can be expected (Figure \ref{figy}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\rel{ours}$. These effects can be seen in Figure \ref{figy} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\rel{ours}$ to
$\rel{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure \ref{figy}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\rel{ours}$. Left: Colors represent $\rel{ours}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\rel{common}$ and $\rel{ours}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{figy}
\end{figure}

@@

For example, although Dorfman pooling improves testing
throughput, we have shown that low values of infection prevalence, or
low values of single test false-negative rates, increase
$\rel{ours}$. These results remain qualitatively similar under varying
parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{figy}). Hence,

@@

>>>>

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern for pools as large as 64 individuals \cite{Lion,
  DorfmanYuvalDor, DilutionCOVID}. Consequently, studies assumed that
the probability of a true-positive (the test's \emph{sensitivity})
does not depend on the number of infected samples in the pool, but
rather on the existence of at least one such sample. Thus, the
probability of a positive result in a pooled test has been (assumed)
identical for a pool with one sample from an infected individual and,
e.g., five such samples. This assumption is common in the group
testing literature \cite{Kim, OptimalDorfmanPool}, as well as in more
specific, COVID-19 focused studies \cite{Simplistic1, Simplistic2}. In
this study we challenge these commonly made assumptions and show how
using a more accurate probabilistic model affects estimation of
false-negative rates for Dorfman pooling.

@@

\subsection{Scheme efficiency}
The efficiency of Dorfman pooling scheme is the expected number of
tests utilized. Let $T$ the number of tests. Then:
\begin{align}
  \begin{split}
    \mathbb{E}[T] &= 1 + N\Pr(\data=1)\\
    %
    %
    %
    &= 1 + N\sum_{\theta} \Pr(\theta) \Pr(\data=1|\theta) \\
    %
    %
    %
    &= 1 + N\sum_{k=0}^N \Pr(\sum \theta_i = k) \Pr(\data=1|\sum \theta_i = k) \\
    %
    %
    %
    &= 1 + N\sum_{k=0}^N \binom{N}{k}q^k(1-q)^k (1-\Sp(1-\Se)^k) \\
    %
    %
    %
    &= 1+N - N\Sp\sum_{k=0}^N \binom{N}{k}q^k(1-q)^k (1-\Se)^k \\
    %
    %
    %
    %&= 1+N - N\Sp \sum_{k=0}^N \binom{N}{k}(q-q\Se)^k(1-q)^k \\
    %
    %
    %
    &= 1+N - N\Sp (1-q\Se)^N\\    
  \end{split}
\end{align}

Compare to the common assumption:

\begin{align}
  \begin{split}
    \mathbb{E}[T] &= 1 + N\Pr(\data=1)\\
    %
    %
    %
    &= 1 + N\sum_{\theta} \Pr(\theta) \Pr(\data=1|\theta) \\
    %
    %
    %
    &= 1 + N\Big[\Pr(\theta_i = 0,\ \forall i)(1-\Sp) + \Pr(\exists i\ \theta_i =1)\Se\Big] \\
    %
    %
    %
    &= 1 + N\Big [(1-q)^N(1-\Sp) + (1-(1-q)^N)\Se \Big ] \\
    %
    %
    %
    &= 1 + N\Big [(1-q)^N(1-\Sp - \Se) + \Se \Big ] \\
    %
    %
    %
    &= 1 + \Se N + N(1-q)^N(1-\Sp - \Se).
  \end{split}
\end{align}

The difference per person is:
\begin{equation}
  1 - \Sp(1-q\Se)^N - \Se - (1-q)^N(1-\Sp - \Se)
\end{equation}

\subsection{Scheme false-negative rate}
We calculate the false-negative rate of Dorfman's pooling scheme.  The
false-negative rate for a single \emph{infected} individual,
henceforth referred to as "Donald" under our model and under the
common assumption. We distinguish three types of false-negative events
when performing pooling. A \emph{single test}'s false-negative is the
event of a negative result upon testing Donald separately, i.e., in an
RT-PCR test without pooling. A \emph{pooled} false-negative occurs
when a pooled test containing Donald's sample (and other samples)
yields a negative result, i.e., the pooling fails to detect at least
one positive result. Lastly, a \emph{scheme} false-negative occurs
when an entire pooling scheme fails to identify Donald as
infected. Our first task is to calculate Dorfman's scheme
false-negative rate. Rephrasing, we wish to answer the following
question: what is the probability of not identifying Donald as
infected under a Dorfman pooling scheme?

Denote the prevalence of infection in the (tested) population $q$. We
assign Donald as individual $1$, so that $\theta_1=1$. Then:

@@

\subsubsection{Comparison metric}
The single test false-negative rate $1-\Se$ and scheme false-negative
rate $\mi$ are compared via:
\begin{equation}\label{eq:erel}
\err{exact} := \frac{\mi - (1-\Se)}{1-\Se} \cdot 100.
\end{equation}
$\err{exact}$ is the percentage increase in the pooling scheme false-negative
rate, relative to the single test false-negative rate.

According to the common assumption, the scheme false-negative rate is
$1-\Se^2$. A short calculation shows that this approximation implies
the percentage increase in scheme false-negative rate is just
$\err{common} := 100\cdot \Se$.

\section{Results}\label{section:results}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, an increase of at least $60\%$ in
$\err{exact}$ can be expected (Figure \ref{figy}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure \ref{figy} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure \ref{figy}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\err{common}$ and $\err{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{figy}
\end{figure}

The efficiency is words...

\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap_efficiency.jpg}
  \caption{Relative increase in Dorfman pooling efficiency
    $\eff{exact}$. Left: Colors represent $\eff{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\eff{common}$ and $\eff{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{figy}
\end{figure}

@@

For example, although Dorfman pooling improves testing
throughput, we have shown that low values of infection prevalence, or
low values of single test false-negative rates, increase
$\err{exact}$. These results remain qualitatively similar under varying
parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{figy}). Hence,

@@

%% Todos
%% add refs to dilution and paragraph in start
